Ricci curvature of the rooted phylogenetic subtree-prune-regraft graph

### 2015-02-06 curvature.tex
degree change lemmas and proofs

%---[ Distance measures, invocations, etc ]-------------------------------------
\newattribute{dspr}{d_{\mathrm{SPR}}}

@@

@@

>>>>

%---[ Distance measures, invocations, etc ]-------------------------------------
\newattribute{dspr}{d_{\mathrm{SPR}}}
\newattribute{degree}{deg}

@@

Song (2003) proved the following:

\begin{lemma}
	\label{lem:degree_extremes}
	Let $T$ be a rooted phylogenetic tree with $n$ leaves. Then
	\begin{enumerate}
		\item $\degree{T} = 3n^2 - 13n + 14$, if $T$ is a ladder tree\todo{notation? Song used $\lambda$-balanced},
		\item $\degree{T} = 4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}$, if $T$ is a balanced tree, and
		\item  $3n^2 - 13n + 14 \le \degree{T} \le 4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}$, otherwise.
	\end{enumerate}
\end{lemma}

We now bound the relative and exact difference in SPR degree between two trees with $n$ leaves.

\begin{lemma}
	\label{lem:degree_max_delta}
	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, and assume without loss of generality that $\degree{T} \le \degree{S}$.
	Then:
	\begin{enumerate}
		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{3}{4}$, and
		\item $\degree{S} - \degree{T} \le n^2 - 5n + 6$.
	\end{enumerate}
\end{lemma}
\begin{proof}
	We first prove (1).
	Clearly $\degree{T} / \degree{S}$ is minimized when $T$ is of minimum degree and $S$ of maximum degree.
	Then, by Lemma~\ref{lem:degree_extremes}, $\degree{T} = 3n^2 - 13n + 14$ and $\degree{S} = 4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}$.
	We thus have:
	\begin{align}
		\frac{\degree{T}}{\degree{S}} \ge\ &\frac{3n^2 - 13n + 14}{4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4(n-2)^2 - 2(n-2)} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 16n + 16 - 2(n-2)} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 18n + 20} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 17\frac{1}{3}n + 18} &\forall n \ge 3 \\
		\ge\ &\frac{3}{4} &\forall n \ge 3
	\end{align}

	We now prove (2).
	$\degree{S} - \degree{T}$  is maximized when $T$ and $S$ are of minimum and maximum degree.
	We thus have:
	\begin{align}
		\degree{S} - \degree{T}
		\le\ &(4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)})
		- (3n^2 - 13n + 14) \\
		\le\ &(4(n-2)^2 - 2(n-2)) - (3n^2 - 13n + 14) \\
		\le\ &4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\
		\le\ &n^2 - 5n + 6
	\end{align}
\end{proof}

@@

We can now use Lemma~\ref{lem:degree_change} to improve on our bounds in Lemma~\ref{lem:degree_max_delta} for the case of two adjacent trees.

\begin{lemma}
	\label{lem:degree_max_delta_adjacent}
	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, such that $\degree{T} \le \degree{S}$ and $\dspr{T,S} = 1$.
	Then:
	\begin{enumerate}
		\item $\degree{S} - \degree{T} \le 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2 $.
		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{5}{6}$, $\forall n \ge 4$, and
		\item $\lim_{n\rightarrow\infty}\frac{\degree{T}}{\degree{S}} =  \frac{6}{7}$.
	\end{enumerate}
\end{lemma}
\begin{proof}
	We first prove (1).
	By Lemma~\ref{lem:degree_change}, $\degree{S} - \degree{T} = 2(k(a-b) + i - j)$.
	This value is maximized by minimizing $b$, namely by setting $b=0$.
	The resulting equation $2(ka + i - j)$ is similarly maximized by setting $i=1$ and balancing the sides of the square $ka$ with as many leaves as possible.
	There are two cases, depending on whether the subtree of $k$ leaves is moved to an ancestor or not. If not, then we set $j=1$ and balance the square with the remaining $n-b-i-j = n-2$ leaves
	Thus, $\degree{S} - \degree{T} \le 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2 $.
	Note that this corresponds to moving the bottom subtree of $\floor{\frac{n-2}{2}}$ or $\ceil{\frac{n-2}{2}}$ leaves in a ladder tree to the rootmost leaf of the tree.

	If the subtree of $k$ leaves is moved to an ancestor, then we gain an additional leaf to balance the square but increase $j$.
	This corresponds to moving the bottom subtree of $\floor{\frac{n}{2}}$ or $\ceil{\frac{n}{2}}$ leaves in a ladder tree to the root.
	Namely, we have $2(ka + 1 - j)$, where $j = n - k = a + 1$.
	If we move the additional leaf, we have:
$\degree{S} - \degree{T} \le 2\ceil{\frac{n}{2}}\floor{\frac{n-2}{2}} -  (\floor{\frac{n-2}{2}} + 1) = 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}}$.
Similarly, if we do not move the additional leaf, we have:
$\degree{S} - \degree{T} \le 2\ceil{\frac{n-2}{2}}\floor{\frac{n}{2}} -  (\ceil{\frac{n-2}{2}} + 1) = 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}}$.

	The relative change in degree, $\frac{\degree{T}}{\degree{S}}$, is maximized when $\degree{T}$ is as small as possible and $\degree{S} - \degree{T}$ is maximal.
	That is, $\degree{S} = 3n^2 - 13n + 14$ and $\degree{S} + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}$.
	We then have:
	\begin{align}
		\frac{\degree{T}}{\degree{S}} &= \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}} \\
		&\ge \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + \frac{1}{2}(n-2)^2} \\
		&\ge \frac{3n^2 - 13n + 14}{3.5n^2 - 15n + 16}
	\end{align}
	Thus (2) and (3) follow.
